Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-12-10 LatticeOptics/submission_rev_sci_inst_revised/latticeoptics.tex
revisions rev sci inst v1

\date{\today}
\begin{abstract}
We describe the design and implementation of a stable high-power \unit{1064}{nm} laser system to generate optical lattices for experiments with ultracold quantum gases. The system is based on a low-noise laser amplified by an array of four heavily modified, high-power fiber amplifiers. The beam intensity is stabilized and controlled with a nonlinear feedback loop.
Using real-time monitoring of the resulting optical lattice, we find the stability of the lattice site positions to be well below the lattice spacing over the course of hours. The pointing stability of the optical lattice beams is around one lattice spacing and the long-term (six month) relative stability of the lattice spacing itself is 0.5\% RMS.
\end{abstract}
\maketitle

@@

\begin{figure*}
  \begin{center}
    \includegraphics{Figure2.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified and free-running Nufern fiber amplifiers and the Mephisto seed, as well as an independent Mephisto MOPA 55W (not used for the optical lattice presented in this manuscript). (b) Shows the open-loop and closed-loop behavior of the high-power feedback loop using the  modified fiber amplifiers, as described in the main text. (c) Shows the open-loop behavior of the low-power feedback loop using the modified fiber amplifiers, including an additional low-pass filtering stage to limit the loop bandwidth at various constant powers. (d) Closed-loop behavior of the low-power feedback loop without additional low-pass filtering, measured on the same setup as panel (c).}\label{fig:noises}
  \end{center}
\end{figure*}

@@

Because two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low-power loop is always manually railed to its high extreme when the high-power loop is active.
This is done by closing a mechanical shutter in front of the photodiode and setting the set-point to a high value, which quickly winds up the integrator to its maximum output.
Conversely, the digital high-power loop is disabled in software when the low-power loop is active.

@@

In contrast, the low-power loop that controls the lattice laser during the \emph{interaction} phase of the experiment is critically important:
\begin{enumerate}
  \item Its effect on the RIN can limit the system temperature and dominate heating rates, limiting the amount of time available for atom interaction~\cite{Blatt2015}.
  \item It needs to avoid varying the beam pointing because the trap position matters during this phase of the experiment.
  \item It must have closed-loop bandwidths $\geq\unit{1}{kHz}$, and open-loop modulation bandwidths $\geq \unit{100}{kHz}$.
\end{enumerate}
To this end, we use an all-custom loop based on a TeO$_2$ AOM (Gooch \& Housego AOMO 3080-198), where the power of the RF supplied to the AOM is used to actuate the fast, low-power feedback loop.
The AOM was chosen for its low thermal lensing and easily-accessible RF powers.
It is mounted on a monolithic, water-cooled flexure mount that allows optimization of the AOM efficiency, while remaining stable over years of operation.

\begin{figure}
  \begin{center}
    \includegraphics{Figure6.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in Fig.~\ref{fig:circuits}(c). As seen in Fig.~\ref{fig:optical_layout}, a pair of low-noise photodiodes [Fig.~\ref{fig:circuits}(a)] measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C).
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes.
One of the sampled beams is attenuated by an additional factor of 20 by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers \unit{0-20}{W} of the beam directed at the atoms, and controls the Berek-compensator-based, high-power feedback loop.
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers \unit{0-1}{W}.
To prevent damage to the low-power photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high-power loop is in use.
The low-power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit [Fig.~\ref{fig:circuits}(c)], which compares it to the control system's set-point, generated by a digital-to-analog converter.
The loop filter's output controls the IF port of a microwave mixer [Minicircuits ZFM-2-S+, see Fig.~\ref{fig:circuits}(d)], to modulate the RF power supplied by a low-noise phase-locked loop [PLL, see Fig.~\ref{fig:circuits}(b)].
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB), schematically shown in Fig.~\ref{fig:circuits}(b), that implements a phase locked loop (PLL) based on the ADF4002 IC and a low-phase-noise voltage-controlled oscillator (VCO, Crystek CVCO55CL-0060-0110), preamplified with a low-noise amplifier (Minicircuits PSA4-5043+).
The PLL is locked to the \unit{10}{MHz} frequency reference distributed throughout the lab to ensure absolute frequency stability.
In addition to PLL functionality, the board also contains two digitally-controlled, high-isolation RF switches, giving a total isolation exceeding \unit{90}{dB}.
This device allows us to drive the AOM with a single, clean tone.
The spectrum and relative intensity noise of the PLL can be found in Fig.~\ref{fig:pll}.
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of a typical VCO but without the slow frequency drift characteristic to these devices.
Elimination of the carrier drift is worth the added phase noise because any drift of the frequency affects the pointing of the laser beam passing through the AOM, which would misalign the optical lattice.

\begin{figure}
  \begin{center}
    \includegraphics{Figure7.pdf}
    \caption{\textbf{PLL performance} (a) The spectrum of the PLL, exhibiting phase noise of better than $-90$ dBc/Hz at $100$~kHz away from the carrier. (b) The RIN of the local oscillator based on the PLL. The noise level is seen to be at or below the noise floor of the measurement device.}\label{fig:pll}
  \end{center}
\end{figure}

@@

The chosen photodiode (Hamamatsu G8370-01) is an InGaS based, small-area, low-noise design with a high efficiency and low thermal sensitivity at \unit{1064}{nm}.
Prior to use, the protective window is removed from the photodiode to prevent interference effects, as seen in Fig.~\ref{fig:cut_pd}.
Further, to reduce the capacitance of the system, a reverse bias of \unit{6.9}{V} is applied to the photodiode by a temperature-stabilized voltage reference (Linear Technology LM399H).
The transimpedance circuit is shown in Fig.~\ref{fig:circuits}(a).
It converts the photo current supplied by the photodiode to a voltage signal using a low-noise and high-bandwidth op-amp (Texas Instruments OPA843).
To reduce electrical pickup and prevent ground loops, the electronic assembly is housed in a metal enclosure and mounted to the optical table using electrically insulating mica posts.

>>>>

\date{\today}
\begin{abstract}
We describe the design and implementation of a stable high-power \unit{1064}{nm} laser system to generate optical lattices for experiments with ultracold quantum gases. The system is based on a low-noise laser amplified by an array of four heavily modified, high-power fiber amplifiers. The beam intensity is stabilized and controlled with a nonlinear feedback loop.
Using real-time monitoring of the resulting optical lattice, we find the stability of the lattice site positions to be well below the lattice spacing over the course of hours. The stability of the underlying harmonic trap produced by the optical lattice beams is around one lattice spacing and the long-term (six month) relative RMS stability of the lattice spacing itself is 0.5\%.
\end{abstract}
\maketitle

@@

\begin{figure*}
  \begin{center}
    \includegraphics{Figure2.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified and free-running Nufern fiber amplifiers and the Mephisto seed, as well as an independent Mephisto MOPA 55W (not used for the optical lattice presented in this manuscript). (b) Shows the open-loop and closed-loop behavior of the high-power feedback loop using the  modified fiber amplifiers, as described in the main text. (c) Shows the open-loop behavior of the low-power feedback loop using the modified fiber amplifiers at various constant powers. (d) Closed-loop behavior of the low-power feedback loop (without additional low-pass filtering), measured on the same setup as panel (c).}\label{fig:noises}
  \end{center}
\end{figure*}

@@

Because two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low-power loop is always manually railed to its high extreme when the high-power loop is active.
This is done by closing a mechanical shutter (Stanford Research Systems SR475) in front of the photodiode and setting the set-point to a high value, which quickly winds up the integrator to its maximum output.
Conversely, the digital high-power loop is disabled in software when the low-power loop is active.

@@

In contrast, the low-power loop that controls the lattice laser during the \emph{interaction} phase of the experiment is critically important:
\begin{enumerate}
  \item Its effect on the RIN can limit the system temperature and dominate heating rates, limiting the amount of time available for atom interaction~\cite{Blatt2015}.
  \item It needs to avoid varying the beam pointing because the trap position matters during this phase of the experiment.
  \item It must have closed-loop bandwidths $\geq\unit{1}{kHz}$, and open-loop modulation bandwidths $\geq \unit{100}{kHz}$.
\end{enumerate}
To this end, we use an nonlinear feedback loop based on a TeO$_2$ AOM (Gooch \& Housego AOMO 3080-198), where the power of the RF supplied to the AOM is used to actuate the fast, low-power feedback loop.
The AOM was chosen for its low thermal lensing and easily-accessible RF powers.
It is mounted on a monolithic, water-cooled flexure mount that allows optimization of the AOM efficiency, while remaining stable over years of operation.

\begin{figure}
  \begin{center}
    \includegraphics{Figure6.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in Fig.~\ref{fig:circuits}(c). As seen in Fig.~\ref{fig:optical_layout}, a pair of low-noise photodiodes [Fig.~\ref{fig:circuits}(a)] measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C).
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes.
One of the sampled beams is attenuated by an additional factor of 20 by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers \unit{0-20}{W} of the beam directed at the atoms, and controls the Berek-compensator-based, high-power feedback loop.
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers \unit{0-1}{W}.
%To prevent damage to the low-power photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high-power loop is in use.
The low-power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit [Fig.~\ref{fig:circuits}(c)], which compares it to the control system's set-point, generated by a digital-to-analog converter.
The loop filter's output controls the IF port of a microwave mixer [Minicircuits ZFM-2-S+, see Fig.~\ref{fig:circuits}(d)], to modulate the RF power supplied by a low-noise phase-locked loop [PLL, see Fig.~\ref{fig:circuits}(b)].
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB), schematically shown in Fig.~\ref{fig:circuits}(b), that implements a phase locked loop (PLL) based on the ADF4002 IC and a low-phase-noise voltage-controlled oscillator (VCO, Crystek CVCO55CL-0060-0110), preamplified with a low-noise amplifier (Minicircuits PSA4-5043+).
The PLL is locked to the \unit{10}{MHz} frequency reference distributed throughout the lab to ensure absolute frequency stability.
In addition to PLL functionality, the board also contains two digitally-controlled, high-isolation RF switches, giving a total RF isolation exceeding \unit{90}{dB}.
This device allows us to drive the AOM with a single, clean tone.
%The spectrum and relative intensity noise of the PLL can be found in Fig.~\ref{fig:pll}.
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of a typical VCO but without the slow frequency drift characteristic to these devices.
Elimination of the carrier drift is worth the added phase noise because any drift of the frequency affects the pointing of the laser beam passing through the AOM, which would misalign the optical lattice.

%\begin{figure}
%  \begin{center}
%    \includegraphics{Figure7.pdf}
%    \caption{\textbf{PLL performance} (a) The spectrum of the PLL, exhibiting phase noise of better than $-90$ dBc/Hz at $100$~kHz away from the carrier. (b) The RIN of the local oscillator based on the PLL. The noise level is seen to be at or below the noise floor of the measurement device.}\label{fig:pll}
%  \end{center}
%\end{figure}

@@

The chosen photodiode (Hamamatsu G8370-01) is an InGaAs based, small-area, low-noise design with a high efficiency and low thermal sensitivity at \unit{1064}{nm}.
Prior to use, the protective window is removed from the photodiode to prevent interference effects, as seen in Fig.~\ref{fig:cut_pd}.
Further, to reduce the capacitance of the system, a reverse bias of \unit{6.9}{V} is applied to the photodiode by a temperature-stabilized voltage reference (Linear Technology LM399H).
The transimpedance circuit is shown in Fig.~\ref{fig:circuits}(a).
It converts the photo current supplied by the photodiode to a voltage signal using a low-noise and high-bandwidth op-amp (Texas Instruments OPA843).
To reduce electrical pickup and prevent ground loops, the electronic assembly is housed in a metal enclosure and mounted to the optical table using electrically insulating mica posts.
